### 2021-05-28 submission/manuscript.tex
Finish draft plus edits

Antibiotics are a major risk factor for \emph{Clostridioides difficile}
infections (CDIs) because of their impact on the microbiota. However,
non-antibiotic medications such as the ubiquitous osmotic laxative
polyethylene glycol (PEG) 3350, also alter the microbiota, but whether
PEG impacts CDI susceptibility and clearance is unclear. To examine how
PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day
doses of 15\% PEG in the drinking water and then challenged the mice
with \emph{C. difficile} 630. We used clindamycin-treated mice as a
control because they consistently clear \emph{C. difficile} within 10
days post-challenge (dpc). PEG treatment alone was sufficient to render
mice susceptible and 5-day PEG-treated mice remain colonized for up to
30 dpc. In contrast, 1-day PEG treated mice were transiently colonized,
clearing \emph{C. difficile} within 7 dpc. To examine how PEG treatment
impacts clearance, we administered a 1-day PEG treatment to
clindamycin-treated, \emph{C. difficile}-challenged mice. Administering
PEG to mice after \emph{C. difficile} challenge prolonged colonization
up to 30 dpc. When we trained a random forest model with community data
from 5 dpc, we were able to predict which mice would exhibit prolonged
colonization (AUROC = 0.90). Five of the top ten bacterial features
important for predicting prolonged colonization had high relative
abundances in the community. Examining the dynamics of these bacterial
during the post-challenge period revealed patterns in the relative
abundances of \emph{Bacteroides}, \emph{Enterobacteriaceae},
\emph{Porphyromonadaceae}, \emph{Lachnospiraceae}, and
\emph{Akkermansia} that were associated with prolonged \emph{C.
difficile} colonization in PEG-treated mice.

@@

Diarrheal samples induced by medications such as laxatives are typically
rejected for \emph{Clostridiodes difficile} testing. However, there are
some microbiota similarities between diarrheal and \emph{C. difficile}
infection (CDI) communities such as lower diversity compared to healthy
controls, which lead us to hypothesize that diarrhea may be an indicator
of \emph{C. difficile} risk. We explored how osmotic laxatives disrupt
the microbiota's colonization resistance to \emph{C. difficile} by
administering a laxative to mice either before or after \emph{C.
difficile} challenge Our findings suggest the osmotic laxative disrupts
colonization resistance to \emph{C. difficile}, as well as clearance in
mice already colonized with \emph{C. difficile}. Considering that most
hospitals recommend not performing \emph{C. difficile} testing on
patients taking laxatives and laxatives are used when administering
fecal microbiota transplants via colonoscopy to patients with recurrent
CDIs, further studies are needed to evaluate if laxatives impact
microbiota colonization resistance in humans.

@@

Many of the non-antibiotic medications associated with CDIs are known to
modulate gastrointestinal motility leading to either increased or
decreased colonic transit time, which in turn also strongly impacts
microbiota composition and function (10, 11). Stool consistency often
serves as an approximation of intestinal motility (10). Our group has
shown that when \emph{C. difficile} negative controls are separated into
two groups based on stool consistency, there are shared microbiota
features such as lower alpha diversity in samples from CDI patients and
control patients with diarrhea compared to control samples that were
\emph{C. difficile} negative with non-diarrheal consistency (12). These
results led to a hypothesis that bacterial communities from patients
experiencing diarrhea are susceptible to developing CDIs.

Osmotic laxatives can lead to diarrhea depending on the administered
dose and temporarily disrupt the human intestinal microbiota (13). The
ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in
Miralax, Nulytely, and Golytely and is also commonly used as bowel
preparation for colonoscopies. Interestingly, previous studies have
shown that treating mice with PEG alone altered microbiota composition,
reduced acetate and butyrate production, altered the mucus barrier, and
rendered the mice susceptible to \emph{C. difficile} colonization
(14--17). The mucus barrier is thought to mediate protection from CDIs
by protecting intestinal epithelial cells from the toxins produced by
\emph{C. difficile} (18, 19). Whether laxative administration results in
more severe CDIs in mice and how long mice remain colonized with
\emph{C. difficile} after challenge is unclear.

@@

Our group has used C57BL/6 mice to characterize how antibiotics disrupt
the microbiota and influence \emph{C. difficile} susceptibility and
clearance (24--26). Although, two groups have now shown PEG treatment
alone renders mice susceptible to \emph{C. difficile} (15, 17), these
studies have raised additional questions regarding the dynamics and
severity of infection as well as the role of laxative treatment in
\emph{C. difficile} clearance. Addressing these questions will better
inform how we think about laxatives in the context of CDIs. Here, we
characterized how long PEG-treated mice remain susceptible, whether PEG
treatment results in more sustained \emph{C. difficile} colonization and
severe CDI than mice treated with clindamycin, and if PEG treatment
after challenge can promote \emph{C. difficile} clearance.

@@

\textbf{5-day laxative treatment led to prolonged \emph{C. difficile}
colonization in mice.} Building off of previous work that showed
treating mice with the osmotic laxative, PEG 3350, rendered mice
susceptible to \emph{C. difficile} colonization (15, 17), we decided to
test how long \emph{C. difficile} colonization is sustained and how long
PEG-treated mice remain susceptible to \emph{C. difficile}. We compared
three groups of mice treated with PEG 3350 to one group of mice treated
with our standard 10 mg/kg clindamycin treatment, which temporarily
renders the mice susceptible to \emph{C. difficile} colonization, with
mice typically clearing \emph{C. difficile} within 10 days
post-challenge (9, 26). All three groups of PEG-treated mice were
administered a 15\% PEG solution in the drinking water for 5-days: 1.
received no additional treatment, 2. was also treated with clindamycin,
and 3. was allowed to recover for 10 days prior to challenge (Fig. 1A).
PEG treatment resulted in weight loss in all 3 groups of PEG-treated
mice relative to their baseline weights, with the greatest change in
weight observed on the fifth day of PEG treatment. The mice recovered
most of the lost weight by five days after treatment (Fig. 1B). After
either the PEG, clindamycin, or PEG and clindamycin treatment all mice
were challenged with 10\textsuperscript{5} \emph{C. difficile} 630
spores (Fig. 1A). All treatments rendered mice susceptible to \emph{C.
difficile} colonization. However, PEG-treated mice remained colonized
with \emph{C. difficile} at a high level through thirty days
post-challenge (Fig. 1C). In contrast, the clindamycin-treated mice
cleared \emph{C. difficile} within ten days post-challenge (Fig. 1C).
Therefore, PEG treatment led to sustained colonization in contrast to
clindamycin mice that naturally cleared \emph{C. difficile} within ten
days post-challenge.

Notably, we also found PEG-treated mice were still susceptible to
\emph{C. difficile} colonization after a 10-day recovery period,
although \emph{C. difficile} was not detectable in most of the group in
the initial five days post-challenge (Fig. 1C, S1A). One mouse was found
dead on the 6th day post-challenge, presumably due to \emph{C.
difficile} as the bacteria became detectable in stool samples from that
mouse on the 4th day post-challenge (Fig. S1A, mouse 10). From 8 days
post-challenge onward, the density of \emph{C. difficile} stabilized in
the 10-day recovery group and remained high through 20-30 days
post-challenge (Fig. 1C). Thus, osmotic laxative treatment alone was
sufficient to render mice susceptible to prolonged \emph{C. difficile}
colonization and PEG-treated mice remained susceptible through ten days
post-treatment.

\textbf{5-day laxative treatment differentially disrupted the fecal
microbiota compared to clindamycin treatment.} Since osmotic laxatives
and clindamycin have previously been shown to disrupt the murine
microbiota (14--17), we hypothesized the different \emph{C. difficile}
colonization dynamics between mice treated with the osmotic laxative or
clindamycin were due to the two drugs having differential effects on the
microbiota. We profiled the stool microbiota over time by sequencing the
V4 region of the 16S rRNA gene to compare changes across treatment
groups. We found time and treatment group explained half of the observed
variation between fecal communities with most of the remaining variation
explained by interactions between treatment group and other experimental
variables including time, cage effects, and sequencing preparation plate
(PERMANOVA combined R\textsuperscript{2} = 0.95, \emph{P} \textless{}
0.001, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the
well-documented phenomenon that mice housed in the same cages have
similar microbial communities due to coprophagy (27). We tried to
minimize the impact of cage effects on our experiment by breaking up
littermates when assigning mice to treatment groups. Importantly,
although we conducted a total of 5 separate experiments, the experiment
number and its interaction with treatment group did no significantly
explain the observed variation in fecal communities (Data Set S1, Sheet
X). None of the treatment groups recovered to their baseline community
structure either 10 or 30 days post-challenge suggesting other community
features besides recovery to baseline were responsible for the prolonged
\emph{C. difficile} colonization in PEG-treated mice (Fig. 2B).

Since none of the communities completely recovered in the follow-up
period after treatments, we next profiled community diversity and
composition. We examined the alpha diversity dynamics by calculating the
communities' Shannon diversity. Although both clindamycin and PEG
treatments decreased diversity, Shannon diversity was lower in the
groups of mice that received PEG treatment compared to those that
received clindamycin through thirty days post-challenge (Fig. 2C). We
next examined the bacterial genera that shifted after PEG treatment by
comparing the baseline samples of mice treated with only PEG to samples
from the same mice one day post-PEG-treatment. We found 18 genera that
were altered by PEG treatment (Data Set S1, Sheet X). The majority of
the bacterial relative abundances decreased after the PEG treatment, but
the relative abundance among members of the \emph{Enterobacteriaceae}
and \emph{Bacteroides} increased. The increase in \emph{Bacteroides}
relative abundance was unique to PEG treated mice, as the
\emph{Bacteroides} relative abundance actually decreased in clindamycin
treated mice (Fig. 2D). Finally, we identified the genera whose relative
abundance differed across treatment groups over multiple time points. Of
the 33 genera that were different between treatment groups, 24 genera
were different over multiple time points (Fig. 2E, Data Set S1, Sheet
X). Thus, PEG had a significant impact on the fecal microbiota that was
maintained over time and was distinct from clindamycin treatment.

Interestingly, \emph{C. difficile} was not immediately detectable in the
stools of the PEG-treated mice that were allowed to recover for 10 days
prior to challenge. We decided to examine if there were genera that
changed during the post-challenge period when the group median \emph{C.
difficile} shifted from undetectable at 1 day post-challenge to
detectable in the stool samples with the density stabilizing around 8
days post-challenge (Fig. S1A). We found no bacteria with relative
abundances that were significantly different over the two time points
after multiple hypothesis correction (Data Set S1, Sheet X). However,
there was also wide variation between individual mice regarding when
\emph{C. difficile} became detectable (Fig. S1A) as well as the relative
abundances of bacterial genera in the communities (Fig. S1B). For
example, two mice had a high relative abundance of
\emph{Enterobacteriaceae} throughout the post-challenge period and this
corresponded to mouse 10, which died on the sixth day post-challenge and
mouse 11, where \emph{C. difficile} was present at a high density from
the 4th day post-challenge onward (Fig. S1B). While we did not identify
a clear signal to explain the delayed appearance of \emph{C. difficile}
in the 5-day PEG mice that were allowed to recover for 10 days prior to
challenge, the delay is striking and could reflect changes in microbial
activity or metabolites that were not examined in this study.

\textbf{5-day laxative treatment did not promote more severe CDIs
despite altering the mucosal microbiota.} Given the findings from a
previous study that demonstrated PEG treatment disrupts the mucus layer
and alters the immune response in mice (16), we decided to examine the
impact of PEG treatment on the mucosal microbiota and CDI severity. To
evaluate the mucosal microbiota, we sequenced snips of tissue collected
from the cecum, proximal colon, and distal colon. Similar to what was
observed with the stool samples, alpha diversity was lower in the
PEG-treated mice compared to clindamycin treated mice (Fig. 3A). Alpha
diversity continued to increase over time based on the tissues from
PEG-treated mice collected at 20 and 30 days post-challenge (Fig. 3A,
Data Set S1, Sheet X). Group, time point, and their interactions with
other variables (cage, experiment number, and sample type) explained the
majority of the variation observed in mucosal communities (PERMANOVA
combined R\textsuperscript{2} = 0.83, \emph{P} \textless{} 0.05, Fig.
3B, Data Set S1, Sheet X). We saw the greatest difference in the
relative abundance of the mucosal microbiota between treatment groups
(clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6 days
post-challenge with 10 genera that were significantly different
(\emph{P} \textless{} 0.05) in all three of the tissue types we
collected (cecum, proximal colon, and distal colon; Fig. S2A, Data Set
S1, Sheet X). Interestingly, \emph{Peptostreptococcaceae} (the family
with a sequence that matches \emph{C. difficile}) was one of the genera
that had a significant difference in relative abundance between
treatment groups at 6 days post-challenge. This population was primarily
only present in the 5-day PEG treatment group of mice and decreased in
the proximal and distal colon tissues over time (Fig. S2B). By 30 days
post-challenge, only the relative abundances of \emph{Bacteroides},
\emph{Clostridiales}, \emph{Firmicutes}, and \emph{Ruminococcaceae} were
different between treatment groups and only in the cecum tissues (Fig.
3C, Fig. 2E, Data Set S1, Sheet X). Thus, PEG treatment had a
significant impact on the mucosal microbiota and we detected \emph{C.
difficile} sequences in the cecum, proximal colon, and distal colon
tissue communities.

Because there were differences in the mucosal microbiota including
detectable \emph{C. difficile} sequences in tissues from PEG-treated
mice relative to mice treated with clindamycin, we next examined the
severity of \emph{C. difficile} challenge by evaluating cecum and colon
H\&E stained histopathology (28). However, we found there was no
difference in cecum and colon scores between clindamycin and PEG-treated
mice that were challenged with \emph{C. difficile} at 4 days
post-challenge (Fig. 3D), the time point typically examined in \emph{C.
difficile} 630 challenged mice (29). We also looked at 6 days
post-challenge because that was when there was a large difference in
\emph{C. difficile} density between PEG- and clindamycin-treated mice
(Fig. 1C). Although, there was a slight difference in the colon between
PEG and clindamycin-treated mice, there was not a signifant difference
in the cecum and the overall score was relatively low (1.5-2.5 out of
12, Fig. 3E). Therefore, although PEG treatment had a disruptive effect
on the mucosal microbiota, the impact of \emph{C. difficile} 630
challenge on the cecum and colon was similar between PEG and clindamycin
treated mice.

\textbf{\emph{C. difficile} challenge did not have a synergistic
disruptive effect on the microbiota of PEG-treated mice.} Because
\emph{C. difficile} itself can have an impact on the microbiota (30), we
also sequenced the tissue and stools of mock-challenged clindamycin and
5-day PEG treated mice. Examining the stools of the mock-challenged mice
revealed similar bacterial disruptions as the \emph{C. difficile}
challenged mice (Fig. S3A-C). Similarly, there was no difference between
the tissues of mock and \emph{C. difficile} challenged mice (Fig.
S3D-F). Thus, most of the microbiota alterations we observed in the
PEG-treated mice were a result of the laxative and not an interaction
between the laxative and \emph{C. difficile}.

\textbf{1-day laxative treatment resulted in transient \emph{C.
difficile} colonization and minor microbiota disruption.} Next, we
examined how a shorter osmotic laxative perturbation would impact the
microbiome and susceptibility to \emph{C. difficile}. We administered
either a 1-day PEG treatment, a 1-day PEG treatment with a 1-day
recovery period, or clindamycin to mice before challenging them with
\emph{C. difficile} (Fig. 3A). In contrast to the 5-day PEG treated
mice, the 1-day PEG treated mice were only transiently colonized and
cleared \emph{C. difficile} by 7 days post-challenge (Fig. 3B). The
stool communities of PEG-treated mice were also only transiently
disrupted, with Shannon diversity recovering by 7 days post-challenge
(Fig. 3C-D). We found the relative abundances of 14 genera were impacted
by treatment, but recovered close to baseline levels by 7 days
post-challenge including \emph{Enterobacteriaceae},
\emph{Clostridiales}, \emph{Porpyromonadaceae}, and
\emph{Ruminococcaceae} (Fig. 3E, Data Set S1, Sheet X). These findings
suggest the duration of the PEG treatment was relevant, with shorter
treatments resulting in a transient loss of \emph{C. difficile}
colonization resistance.

\textbf{Post-challenge laxative treatment disrupted clearance in
clindamycin-treated mice regardless of whether an FMT was also
administered.} Since a 1-day PEG treatment resulted in a more mild
microbiota perturbation, we decided to use the 1-day treatment to
examine the hypothesis that PEG helps to flush \emph{C. difficile}
spores from the intestine. To examine the impact of PEG treatment on
\emph{C. difficile} clearance, we treated 4 groups of mice with
clindamycin and then challenged all mice with \emph{C. difficile} before
administering the following treatments: no additional treatment, 1-day
PEG immediately after challenge, and 1-day PEG treatment 3 days after
challenge followed by either administration of an FMT or PBS solution by
oral gavage (Fig. 5A). Contrary to our hypothesis, all groups of mice
that received PEG exhibited prolonged \emph{C. difficile} colonization
(Fig. 5B).

>>>>

Antibiotics are a major risk factor for \emph{Clostridioides difficile}
infections (CDIs) because of their impact on the microbiota. However,
non-antibiotic medications such as the ubiquitous osmotic laxative
polyethylene glycol (PEG) 3350, also alter the microbiota, but whether
PEG impacts CDI susceptibility and clearance is unclear. To examine how
PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day
doses of 15\% PEG in the drinking water and then challenged the mice
with \emph{C. difficile} 630. We used clindamycin-treated mice as a
control because they consistently clear \emph{C. difficile} within 10
days post-challenge (dpc). PEG treatment alone was sufficient to render
mice susceptible and 5-day PEG-treated mice remain colonized for up to
30 dpc. In contrast, 1-day PEG treated mice were transiently colonized,
clearing \emph{C. difficile} within 7 dpc. To examine how PEG treatment
impacts clearance, we administered a 1-day PEG treatment to
clindamycin-treated, \emph{C. difficile}-challenged mice. Administering
PEG to mice after \emph{C. difficile} challenge prolonged colonization
up to 30 dpc. When we trained a random forest model with community data
from 5 dpc, we were able to predict which mice would exhibit prolonged
colonization (AUROC = 0.90). Examining the dynamics of these bacterial
during the post-challenge period revealed patterns in the relative
abundances of \emph{Bacteroides}, \emph{Enterobacteriaceae},
\emph{Porphyromonadaceae}, \emph{Lachnospiraceae}, and
\emph{Akkermansia} that were associated with prolonged \emph{C.
difficile} colonization in PEG-treated mice.

@@

Diarrheal samples induced by medications such as laxatives are typically
rejected for \emph{Clostridiodes difficile} testing. However, there are
some microbiota similarities between diarrheal and \emph{C. difficile}
infection (CDI) communities such as lower diversity compared to samples
from healthy patients, which lead us to hypothesize that diarrhea may be
an indicator of \emph{C. difficile} risk. We explored how osmotic
laxatives disrupt the microbiota's colonization resistance to \emph{C.
difficile} by administering a laxative to mice either before or after
\emph{C. difficile} challenge. Our findings suggest the osmotic laxative
disrupts colonization resistance to \emph{C. difficile}, as well as
clearance in mice already colonized with \emph{C. difficile}.
Considering that most hospitals recommend not performing \emph{C.
difficile} testing on patients taking laxatives and laxatives are used
when administering fecal microbiota transplants via colonoscopy to
patients with recurrent CDIs, further studies are needed to evaluate if
laxatives impact microbiota colonization resistance in humans.

@@

Many of the non-antibiotic medications associated with CDIs are known to
modulate gastrointestinal motility leading to either increased or
decreased colonic transit time, which in turn also strongly impacts
microbiota composition and function (10, 11). Stool consistency often
serves as an approximation of intestinal motility (10). Our group has
shown that when \emph{C. difficile} negative samples from patients are
separated into two groups based on stool consistency, there are shared
microbiota features such as lower alpha diversity in samples from CDI
patients and \emph{C. difficile} negative patients with diarrhea
compared to samples that were \emph{C. difficile} negative with
non-diarrheal stool consistency (12). These results led to the
hypothesis that bacterial communities from patients experiencing
diarrhea are susceptible to developing CDIs, regardless of how they
developed diarrhea.

Osmotic laxatives can lead to diarrhea depending on the dose
administered and temporarily disrupt the human intestinal microbiota
(13). The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is
found in Miralax, Nulytely, and Golytely and is also commonly used as
bowel preparation for colonoscopies. Interestingly, previous studies
have shown that treating mice with PEG alone altered microbiota
composition, reduced acetate and butyrate production, altered the mucus
barrier, and rendered the mice susceptible to \emph{C. difficile}
colonization (14--17). The mucus barrier is thought to mediate
protection from CDIs by protecting intestinal epithelial cells from the
toxins produced by \emph{C. difficile} (18, 19). Whether laxative
administration results in more severe CDIs in mice and how long mice
remain colonized with \emph{C. difficile} after challenge is unclear.

@@

Our group has used C57BL/6 mice to characterize how antibiotics disrupt
the microbiota and influence \emph{C. difficile} susceptibility and
clearance (24--26). Although two groups have now shown PEG treatment
alone renders mice susceptible to \emph{C. difficile} (15, 17), these
studies have raised additional questions regarding the dynamics and
severity of infection as well as the role of laxative treatment in
\emph{C. difficile} clearance. Here, we characterized how long
PEG-treated mice remain susceptible, whether PEG treatment results in
more sustained \emph{C. difficile} colonization and severe CDI than mice
treated with clindamycin, and if PEG treatment after challenge can
promote \emph{C. difficile} clearance. Addressing these questions will
better inform how we think about laxatives and diarrhea in the context
of CDIs.

@@

\textbf{5-day laxative treatment led to prolonged \emph{C. difficile}
colonization in mice.} Building off of previous work that showed
treating mice with the osmotic laxative, PEG 3350, rendered mice
susceptible to \emph{C. difficile} colonization (15, 17), we decided to
test how long \emph{C. difficile} colonization is sustained and how long
PEG-treated mice remain susceptible to \emph{C. difficile}. We compared
three groups of mice treated with PEG 3350 to one group of mice treated
with our standard 10 mg/kg clindamycin treatment, which temporarily
renders the mice susceptible to \emph{C. difficile} colonization, with
mice typically clearing \emph{C. difficile} within 10 days
post-challenge (9, 26). All three groups of PEG-treated mice were
administered a 15\% PEG solution in the drinking water for 5-days. The
first group received no additional treatment. The second group was also
treated with clindamycin. A third group was allowed to recover for 10
days prior to challenge (Fig. 1A). The PEG treatment resulted in weight
loss for the 3 groups of mice, with the greatest change in weight
observed on the fifth day of the PEG treatment. The mice recovered most
of the lost weight by five days after treatment (Fig. 1B). After either
the PEG, clindamycin, or PEG and clindamycin treatment all mice were
challenged with 10\textsuperscript{5} \emph{C. difficile} 630 spores
(Fig. 1A). All treatments rendered mice susceptible to \emph{C.
difficile} colonization. In contrast to the mice that only received
clindamycin, PEG-treated mice remained colonized with \emph{C.
difficile} at a high level through thirty days post-challenge (Fig. 1C).
The clindamycin-treated mice cleared \emph{C. difficile} within ten days
post-challenge (Fig. 1C). It was noteworthy that PEG-treated mice were
still susceptible to \emph{C. difficile} colonization after a 10-day
recovery period, although \emph{C. difficile} was not detectable in most
of the group in the initial five days post-challenge (Fig. 1C, S1A). One
mouse was found dead on the 6th day post-challenge, presumably due to
\emph{C. difficile} as the bacterium became detectable in stool samples
from that mouse on the 4th day post-challenge (Fig. S1A, mouse 10). From
8 days post-challenge onward, the density of \emph{C. difficile}
stabilized in the 10-day recovery group and remained high through 20-30
days post-challenge (Fig. 1C). Thus, osmotic laxative treatment alone
was sufficient to render mice susceptible to prolonged \emph{C.
difficile} colonization and PEG-treated mice remained susceptible
through ten days post-treatment.

\textbf{5-day laxative treatment differentially disrupted the fecal
microbiota compared to clindamycin treatment.} Since osmotic laxatives
and clindamycin have previously been shown to disrupt the murine
microbiota (14--17), we hypothesized the different \emph{C. difficile}
colonization dynamics between mice treated with the osmotic laxative or
clindamycin were due to the two drugs having differential effects on the
microbiota. We profiled the stool microbiota over time by sequencing the
V4 region of the 16S rRNA gene to compare changes across treatment
groups. We found time (R\textsuperscript{2} = 0.29) and treatment group
(R\textsuperscript{2} = 0.21) explained half of the observed variation
between fecal communities with most of the remaining variation explained
by interactions between treatment group and other experimental variables
including time, cage, and sequencing preparation plate (PERMANOVA
combined R\textsuperscript{2} = 0.95, \emph{P} \textless{} 0.001, Fig.
2A, Data Set S1, sheet 1). None of the treatment groups recovered to
their baseline community structure either 10 or 30 days post-challenge
suggesting other community features besides recovery to baseline were
responsible for the prolonged \emph{C. difficile} colonization in
PEG-treated mice (Fig. 2B).

Since none of the communities completely recovered in the follow-up
period after treatments, we next profiled community diversity and
composition. We examined the alpha diversity dynamics by calculating the
communities' Shannon diversity. Although both clindamycin and PEG
treatments decreased diversity, Shannon diversity was lower in the
groups of mice that received PEG treatment compared to those that
received clindamycin through thirty days post-challenge (Fig. 2C; Data
Set S1, sheet 2). We next examined the bacterial genera that shifted
after PEG treatment by comparing the baseline samples of mice treated
with only PEG to samples from the same mice one day post-PEG-treatment.
We found 18 genera that were altered by PEG treatment (Data Set S1,
sheet 3). The majority of the bacterial relative abundances decreased
after the PEG treatment, but the relative abundance among members of the
\emph{Enterobacteriaceae} and \emph{Bacteroides} increased. The increase
in \emph{Bacteroides} relative abundance was unique to PEG treated mice,
as the \emph{Bacteroides} relative abundance actually decreased in
clindamycin treated mice (Fig. 2D). Finally, we identified the genera
whose relative abundance differed across treatment groups over multiple
time points. Of the 33 genera that were different between treatment
groups, 24 genera were different over multiple time points (Fig. 2E,
Data Set S1, sheet 4). Thus, PEG had a significant impact on the fecal
microbiota that was maintained over time and was distinct from
clindamycin treatment.

Because \emph{C. difficile} was not immediately detectable in the stools
of the PEG-treated mice that were allowed to recover for 10 days prior
to challenge, we decided to examine if there were genera that changed
during the post-challenge period. We compared the communities from when
\emph{C. difficile} shifted from undetectable at 1 day post-challenge to
detectable in the stool samples with the density stabilizing around 8
days post-challenge (Fig. S1A). We found no genera with relative
abundances that were significantly different over the two time points
(Data Set S1, sheet 5). However, there was also wide variation between
individual mice regarding when \emph{C. difficile} became detectable
(Fig. S1A) as well as the relative abundances of bacterial genera in the
communities (Fig. S1B). For example, two mice had a high relative
abundance of \emph{Enterobacteriaceae} throughout the post-challenge
period. One mouse died on the sixth day post-challenge and in the other
\emph{C. difficile} was present at a high density from the 4th day
post-challenge onward (Fig. S1B). While we did not identify a clear
signal to explain the delayed appearance of \emph{C. difficile} in the
5-day PEG mice that were allowed to recover for 10 days prior to
challenge, the delay was striking and could reflect changes in microbial
activity or metabolites that were not examined in this study.

\textbf{5-day laxative treatment did not promote more severe CDIs
despite altering the mucosal microbiota.} Given the findings from a
previous study that demonstrated PEG treatment disrupts the mucus layer
and alters the immune response in mice (16), we decided to examine the
impact of PEG treatment on the mucosal microbiota and CDI severity. To
evaluate the mucosal microbiota, we sequenced communities associated
with tissues collected from the cecum, proximal colon, and distal colon.
Similar to what was observed with the stool samples, the alpha diversity
was lower in the PEG-treated mice compared to clindamycin treated mice
(Fig. 3A, Data Set S1, sheet 6). The alpha diversity of the
tissue-associated community increased in PEG-treated mice collected at
20 and 30 days post-challenge (Fig. 3A). Group (R\textsuperscript{2} =
0.33), time point (R\textsuperscript{2} = 0.11), and their interactions
with other variables (cage, experiment number, and sample type)
explained the majority of the variation observed in mucosal communities
(PERMANOVA combined R\textsuperscript{2} = 0.83, \emph{P} \textless{}
0.05, Fig. 3B, Data Set S1, sheet 7). We saw the greatest difference in
the relative abundance of the mucosal microbiota between treatment
groups (clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6
days post-challenge with 10 genera that were significantly different
(\emph{P} \textless{} 0.05) in all three of the tissue types we
collected (cecum, proximal colon, and distal colon; Fig. S2A, Data Set
S1, sheet 8, 9, and 10). Interestingly, \emph{Peptostreptococcaceae}
(the family with a sequence that matches \emph{C. difficile}) was one of
the genera that had a significant difference in relative abundance
between treatment groups at 6 days post-challenge. This population was
primarily only present in the 5-day PEG treatment group of mice and
decreased in the proximal and distal colon tissues over time (Fig. S2B).
By 30 days post-challenge, only the relative abundances of
\emph{Bacteroides}, \emph{Clostridiales}, \emph{Firmicutes}, and
\emph{Ruminococcaceae} were different between treatment groups and only
in the cecum tissues (Fig. 3C, Fig. 2E, Data Set S1, sheet 8). Thus, PEG
treatment had a significant impact on the mucosal microbiota and we
detected \emph{C. difficile} sequences in the cecum, proximal colon, and
distal colon tissue communities.

Because there were differences in the mucosal microbiota including
detectable \emph{C. difficile} sequences in tissues from PEG-treated
mice relative to mice treated with clindamycin, we next examined the
severity of \emph{C. difficile} challenge by evaluating cecum and colon
histopathology (27). However, we found there was no difference in cecum
and colon scores between clindamycin and PEG-treated mice that were
challenged with \emph{C. difficile} at 4 days post-challenge (Fig. 3D),
the time point typically examined in \emph{C. difficile} 630 challenged
mice (28). We also looked at 6 days post-challenge because that was when
there was a large difference in \emph{C. difficile} density between PEG-
and clindamycin-treated mice (Fig. 1C). Although there was a slight
difference in the histopathology score of the colon between PEG and
clindamycin-treated mice, there was not a signifant difference in the
cecum and the overall score was relatively low (1.5 to 2.5 out of 12,
Fig. 3E). Therefore, although PEG treatment had a disruptive effect on
the mucosal microbiota, the impact of \emph{C. difficile} 630 challenge
on the cecum and colon was similar between PEG and clindamycin treated
mice.

\textbf{\emph{C. difficile} challenge did not have a synergistic
disruptive effect on the microbiota of PEG-treated mice.} Because
\emph{C. difficile} itself can have an impact on the microbiota (29), we
also sequenced the tissue and stools of mock-challenged mice treated
with clindamycin or PEG. Examining the stools of the mock-challenged
mice revealed similar bacterial disruptions as the \emph{C. difficile}
challenged mice (Fig. S3A-C). Similarly, there was no difference between
the communities found in the tissues of mock and \emph{C. difficile}
challenged mice (Fig. S3D-F). Thus, most of the microbiota alterations
we observed in the PEG-treated mice were a result of the laxative and
not an interaction between the laxative and \emph{C. difficile}.

\textbf{1-day laxative treatment resulted in transient \emph{C.
difficile} colonization and minor microbiota disruption.} Next, we
examined how a shorter osmotic laxative perturbation would impact the
microbiome and susceptibility to \emph{C. difficile}. We administered
either a 1-day PEG treatment, a 1-day PEG treatment with a 1-day
recovery period, or clindamycin to mice before challenging them with
\emph{C. difficile} (Fig. 3A). In contrast to the 5-day PEG treated
mice, the 1-day PEG groups were only transiently colonized and cleared
\emph{C. difficile} by 7 days post-challenge (Fig. 3B). The stool
communities of the 1-day PEG treatment groups were also only transiently
disrupted, with Shannon diversity recovering by 7 days post-challenge
(Fig. 3C-D, Data Set S1, sheets 11 and 12). We found the relative
abundances of 14 genera were impacted by treatment, but recovered close
to baseline levels by 7 days post-challenge including
\emph{Enterobacteriaceae}, \emph{Clostridiales},
\emph{Porpyromonadaceae}, and \emph{Ruminococcaceae} (Fig. 3E, Data Set
S1, sheet 13 and 14). These findings suggest the duration of the PEG
treatment was relevant, with shorter treatments resulting in a transient
loss of \emph{C. difficile} colonization resistance.

\textbf{Post-challenge laxative treatment disrupted clearance in
clindamycin-treated mice regardless of whether an FMT was also
administered.} Since a 1-day PEG treatment resulted in a more mild
perturbation of the microbiota, we decided to use the 1-day treatment to
examine the hypothesis that PEG helps to flush \emph{C. difficile}
spores from the intestine. This hypothesis is proposed in the discussion
section of FMT studies where bowel prep is part of the preparation
undergone by patients receiving FMTs via colonoscopy (20--23). To
examine the impact of PEG treatment on \emph{C. difficile} clearance, we
treated 4 groups of mice with clindamycin and then challenged all mice
with \emph{C. difficile} before administering the following treatments:
no additional treatment, 1-day PEG immediately after challenge, and
1-day PEG treatment 3 days after challenge followed by either
administration of an FMT or PBS solution by oral gavage (Fig. 5A).
Contrary to the hypothesis, all groups of mice that received PEG
exhibited prolonged \emph{C. difficile} colonization (Fig. 5B).

We were also interested in exploring whether PEG might help with
engraftment in the context of FMTs. An FMT was prepared under anaerobic
conditions using stool collected from the same group of mice
pre-treatment representing the baseline community. The FMT appeared to
partially restore Shannon diversity but not richness (Fig. 5C-D, Data
Set S1, sheets 15 and 16). Similarly, we saw some overlap between the
communities of mice that received FMT and the mice treated with only
clindamycin after 5 days post-challenge (Fig. 6A, Data Set S1, sheet
17). The increase in Shannon diversity suggests that the FMT did have an
impact on the microbiota, despite seeing prolonged \emph{C. difficile}
colonization in the FMT treated mice. However, only the relative
abundances of \emph{Bacteroidales} and \emph{Porphyromonadaceae}
consistently differed between the mice that received either an FMT or
PBS gavage (Fig. 6B). Overall, we found the relative abundances of 24
genera were different between treatment groups over multiple time points
(Data Set S1, sheet 18). For example, the relative abundance of
\emph{Akkermansia} was increased and the relative abundances of
\emph{Ruminococcaceae}, \emph{Clostridiales}, \emph{Lachnospiraceae},
and \emph{Oscillibacter} were decreased in mice that received PEG after
\emph{C. difficile} challenge relative to clindamycin treated mice (Fig.
6C). In sum, administering PEG actually prolonged \emph{C. difficile}
colonization, including in mice that received an FMT, which only
restored 2 bacterial genera.
